---
title: "Project1_Analysis"
author: "Kshitij Bhat"
date: "11 October 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if('nycflights13' %in% rownames(installed.packages()) == FALSE){
  install.packages("nycflights13")
}

if ('ggplot2' %in% rownames(installed.packages()) == FALSE){
  install.packages("ggplot2")
}
#install.packages("reshape")
#install.packages("plot3D")
library(plot3D)
```

```{r sampling, include= T}

library(nycflights13)
library("ggplot2")

# lets take out the sample where ther is no NA values in delay and has delay greater than 20 minutes

no_na_data_frame <- flights[!is.na(flights$dep_delay) & flights$dep_delay > 20,]


# lets see which flight has maximum departure data

max_delay_main <- max(flights$dep_delay,na.rm = TRUE)

cat("The maximum delay in flights in 2013 is ",max_delay_main," minutes" )
```


#### Below is the outlier function used from existing sources to compare the data performance with and without outliers.

```{r Outlie, eval=TRUE}
outlierKD <- function(dt, var) { 
     var_name <- eval(substitute(var),eval(dt))
     na1 <- sum(is.na(var_name))
     m1 <- mean(var_name, na.rm = T)
     par(mfrow=c(2, 2), oma=c(0,0,3,0))
     boxplot(var_name, main="With outliers")
     hist(var_name, main="With outliers", xlab=NA, ylab=NA)
     outlier <- boxplot.stats(var_name)$out
     mo <- mean(outlier)
     var_name <- ifelse(var_name %in% outlier, NA, var_name)
     boxplot(var_name, main="Without outliers")
     hist(var_name, main="Without outliers", xlab=NA, ylab=NA)
     title("Outlier Check", outer=TRUE)
     na2 <- sum(is.na(var_name))
     cat("Outliers identified:", na2 - na1, "n")
     cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "n")
     cat("Mean of the outliers:", round(mo, 2), "n")
     m2 <- mean(var_name, na.rm = T)
     cat("Mean without removing outliers:", round(m1, 2), "n")
     cat("Mean if we remove outliers:", round(m2, 2), "n")
     response <- "y"
       #readline(prompt="Do you want to remove outliers and to replace with NA? [yes/no]: ")
     if(response == "y" | response == "yes"){
          dt[as.character(substitute(var))] <- invisible(var_name)
          assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
          cat("Outliers successfully removed", "n")
          return(invisible(dt))
     } else{
          cat("Nothing changed", "n")
          return(invisible(var_name))
     }
}

```

#### A histogram of the delays by the airline carriers

```{r extr_data, include=T}
head(flights)



outlierKD(flights,dep_delay)
hist(flights$dep_delay,xlab = "Delay in minutes",main = " Departure delays")

cat("Lets see how the delays vary for different carriers")

#boxplot(flights$dep_delay[flights$dep_delay>0] ~ flights$carrier[flights$dep_delay>0], main="Departure delays for different carriers",xlab="Carrier",ylab="Delay in minutes")
ggplot(data = flights[!is.na(flights$dep_delay),],aes(x=carrier,y=dep_delay)) +
  geom_boxplot()+
  xlab("Carriers") +
  ylab("Delay ranges") +
  ggtitle("Delay ranges for different carriers") +
  theme( axis.line = element_line(colour = "darkblue", size = 1, linetype = "solid"))
cat("Lets check if carrier and distance contribute in predicting the departure_delay\n")


```

#### A barplot to see how many delays more than 20 minutes occurred each month

```{r perm, include=T}

delay_per_month <- c()
total_flights <- c()
for (i in 1:12)
{
  delay_count <- length(no_na_data_frame$month[no_na_data_frame$month == i])
  total_flights <- append(total_flights,length(flights$month[flights$month == i]))
  delay_per_month <- append(delay_per_month,delay_count)
}


plt <- barplot(delay_per_month,
               names.arg = month.abb,
               ylim = c(0,max(delay_per_month)+2000),
               main=" Number of delays > 20 minutes for all months",
               xlab = "Months",
               ylab = "No of delays > 20 mins",
               las =2
                 )
y <- as.matrix(delay_per_month)

text(plt,y+1000,labels =as.character(delay_per_month))

axis(1,at=delay_per_month,labels = month.abb)


```

#### We can infer from the above barplot that the months of June, July, and December have most number of delays greater than 20 minutes

#### Plot to see percentage of flights are being delayed in different months of a year
```{r perc , include=T}
library(ggplot2)

# List to contain the percentage of total flights delayed each month
delay_perc = round((delay_per_month*100)/total_flights,2)

 ggplot(mapping = aes(1:12,delay_perc)) + 
   geom_line() + geom_point() + 
   scale_x_continuous(breaks=1:12,labels = month.abb) + 
   xlab("Months") + ylab("percentage of flights delayed") + 
   scale_y_continuous(limits = c(1,max(delay_perc)+5)) + 
   theme( axis.line = element_line(colour = "darkblue", size = 1, linetype = "solid")) +
   geom_text(aes(label = delay_perc,y=delay_perc + 2), size = 3)


```

#### Comparing the total flights and delayed flights (> 20 mins) per month

```{r comp_mont, include=T}



groups <- rep(c("Delayed_Flights","Total_Flights"), 12)



#creating a list of pair of delay flight and total flights
yvals <- c()

for (i in 1:12){
  yvals <- append(yvals,delay_per_month[i])
  yvals <- append(yvals,total_flights[i])
  
  
}

comp_df <- data.frame("Months"=rep(month.abb,each=2),"Category"= groups, "Total_Flights" = yvals )

ggplot(data = comp_df,aes(fill= Category, y=Total_Flights,x=Months)) + 
    geom_bar(position="dodge",stat="identity",width =0.8) +
    theme( axis.line = element_line(colour = "darkblue", size = 1, linetype = "solid")) +
   geom_text(aes(label = comp_df$Total_Flights,y=comp_df$Total_Flights + 1000), size = 3)+
  ggtitle("Total flights and delayed flights") +
  ylab(" Number of flights")


```



#### lets get a plot where we would know number of delays versus total delays more than 20 minutes  by different airlines for the year 2013.

```{r comparing, include= T}
library(ggplot2)
# lets take out the list of unique aircraft carriers which have a positive delay

car_list <- unique(no_na_data_frame$carrier)

# create a dataframe which has the count of delays by a carrier

delay_frame <- data.frame("Carrier_Code"=c(),"Carrier_Name"=c(),"Delay_Count"=c(),"Total_Flights"=c())



for (i in 1:length(car_list)){
   delay_frame[i,"Carrier_Code"] <- car_list[i]
   delay_frame[i,"Carrier_Name"] <- airlines$name[airlines$carrier == car_list[i]]
  delay_frame[i,"Delay_Count"] <- length(no_na_data_frame$dep_delay[no_na_data_frame$carrier == car_list[i]])
  delay_frame[i,"Total_Flights"] <- length(flights$flight[flights$carrier == car_list[i]])
  
}

xvals <- c()
yvals<- c()

for (k in 1:nrow(delay_frame)) {
  xvals <- append(xvals,rep(delay_frame$Carrier_Code[k],2))
}

for (j in 1:length(xvals))
{
  if(j%%2 ==0)
  {
    yvals <- append(yvals,delay_frame$Total_Flights[delay_frame$Carrier_Code == xvals[j]])
  }
  else
  {
    yvals <- append(yvals,delay_frame$Delay_Count[delay_frame$Carrier_Code == xvals[j]])
  }
}


groups <- rep(c("Delayed Flights","Total Flights"),nrow(delay_frame))

plot_df <- data.frame("Carrier_Codes"= xvals,"Category"=groups,"Total_Flights"=yvals)


ggplot(data=plot_df, aes(fill=Category,y=Total_Flights,x=Carrier_Codes)) + 
    geom_bar(position="dodge",stat="identity",width =1) +
  ggtitle("Comparing Delayed Flights and Total Flights")

```

#### Hence the table is as below:

```{r tab, include=F}

delay_frame

```



#### During the year 2013, for the flights that departed New York City, we can see that:
#### Approximately 1/4th of the number of flights by ExpressJet Airlines have a delay greater than 20 minutes
#### Approximately 1/5th of the number of flights by Jet Blue Airways have a delay greater than 20 minutes
#### Approximately 1/6th of the number of flights by United Airlines have a delay greater than 20 minutes

#### Note: In the above obseravations, the flight delays are not considered in delay counts when the field value for departure delay is 'NA'.


```{r merging, include=T}


# first lets have a look at planes dataframe

head(planes)

# lets chose the relevant columns: tailnum,manufacturer, speed

plane_df <- subset(planes,select = c("tailnum","manufacturer","speed"))

# lets merge the flights dataframe (with delay > 20 minutes) with planes dataframe

flight_plane_df <- merge.data.frame(no_na_data_frame,plane_df,by ="tailnum")

head(flight_plane_df)

```


#### Lets see if delays are common between a particular set of origin and destination airports


```{r org_dest, include=T}
# 
airport_df <- subset(airports, select = c("faa","name"))


# pick only relevant fields from no_na_data_frame
flight_sub_df <- subset(no_na_data_frame,select = c("carrier","origin","dest","dep_delay","arr_delay"))
flight_airport <- merge.data.frame(flight_sub_df,airport_df,by.x = "dest", by.y = "faa")
head(flight_airport)



```

#### We would want to see if there is a correlation between flight arrival delays and other features of the flight dataset

#### Below are the some of the  linear models created and their summary

```{r lin,include=T}
lm1<-lm(arr_delay ~ flight+ distance + air_time,data = no_na_data_frame)
summary(lm1)
aov(lm1)
lm2<-lm(arr_delay ~ dep_delay + flight+ distance + air_time,data = no_na_data_frame)
summary(lm2)
aov(lm2)
#t-tests
carrier_arrivaldel<-t.test(flight_airport$arr_delay)
carrier_arrivaldel
carrier_depdel<-t.test(flight_airport$dep_delay)
carrier_depdel
arrivaldep<-t.test(flight_airport$dep_delay, flight_airport$arr_delay, paird=TRUE)
arrivaldep
```


#### The purpose of this code is to plot bar graphs showing the percentage of flights that each carrier suffers delays from
```{r carrierPercentages}
tableTotal <- table(c(flights$carrier)) #this just shows all of the flights, delayed or not
barplot(tableTotal, las=2)

#prep work here is the same for all 3
lateDeparture<-subset(flights, flights$dep_delay > 20) #subset the flights data set into the specific delays we want
v3<-c(lateDeparture$carrier) #combine into a vector
tablev3<-table(v3) #put the vector in a table for the barplot later.

lateArrival<-subset(flights, flights$arr_delay > 20)
v4<-c(lateArrival$carrier)
tablev4<-table(v4)

lateBoth<-subset(lateDeparture, lateDeparture$arr_delay > 20)
v5<-c(lateBoth$carrier)
tablev5<-table(v5)


#this portion of the code iterates through, calculates, and updates the tables with the percentages of delays per total flights, instaed of just total flights
v3SumTotal = 0
v4SumTotal = 0
v5SumTotal = 0
for (i in 1:length(tableTotal)) {
  v3SumTotal = v3SumTotal + (tablev3[i] / tableTotal[i])
  tablev3[i] = tablev3[i] / tableTotal[i]
}
barplot(tablev3, las=2) #and then plot it.
for (i in 1:length(tableTotal)) {
  v4SumTotal = v4SumTotal + (tablev4[i] / tableTotal[i])
  tablev4[i] = tablev4[i] / tableTotal[i]
}
barplot(tablev4, las=2)
for (i in 1:length(tableTotal)) {
  v5SumTotal = v5SumTotal + (tablev5[i] / tableTotal[i])
  tablev5[i] = tablev5[i] / tableTotal[i]
}

print(v3SumTotal/length(tableTotal))
print(v4SumTotal/length(tableTotal))
print(v5SumTotal/length(tableTotal))

barplot(tablev5, las=2)

#chisq.test(tablev3, tablev5)
```

#### This code generates a 3D histogram showing side-by-side the frequency of various  delays for each of the carriers.

```{r 3DHist}
y_3d <- lateDeparture$carrier #y axis denotes the carriers
x_3d <- lateDeparture$dep_delay #x axis denotes the departure delay
z_3d <- table(x_3d, y_3d) #combine
histOutput <- hist3D(z=z_3d) #plot onto 3D hist. I spent a long time trying to affix labels with this package, and could not get them to showup, so I applied them manually after some testing to make sure I was labelling correctly.

#This is the same 3D histogram for arrival delay.
y_3d2 <- lateArrival$carrier #y axis denotes the carriers
x_3d2 <- lateArrival$arr_delay #x axis denotes the departure delay
z_3d2 <- table(x_3d2, y_3d2) #combine
histOutput <- hist3D(z=z_3d2)
                     
```